\textbf{Varying coefficients correlated velocity models in complex landscapes with application to narwhal behavioral response to ship exposure and airgun exposure}

### 2024-05-08 Latex/main.tex
new paper version

The diffusion models are discussed in Section 2 and simulation study results are shown.
Then,  in Section 3 we give an overview of the narwhal data available for the analysis and the results of the application of our models to this data. The last section is dedicated to a discussion about these results, and potential extension of the models.

@@

\subsubsection{Solution of the SDE}

@@

\begin{equation*} \left\{
	\begin{array}{l}
		dX(t)=V(t)dt \\
		dV(t)=-A(V(t)-\mu)dt+\frac{2\nu}{\sqrt{\pi \tau}} dW(t) 
	\end{array}
	\right.
	\label{eq: RACVM equation bis}
\end{equation*}
where $X$ is the two-dimensional location and $V$ is the two-dimensional velocity.
In this section, we obtain explicit distributional results about the process $X$. Though this model is considered in several papers, we found no mention of the closed form formulas derived here. In \cite{gurarie_correlated_2017}, only the process $V$ is exhaustively studied, and the classical distributional results about this process are used for estimation and simulation purposes.
A more general formulation is considered by \cite{albertsen_generalizing_2018} in which the diffusion matrix is lower triangular matrix with positive diagonal elements whereas and two distinct autocorrelation parameters are considered in both directions allowing for anisotropic movement.
In this case, \cite{albertsen_generalizing_2018} derives an explicit form for the gaussian transition density of the velocity process, where the covariance matrix is expressed using a Kronecker sum \cite{albertsen_generalizing_2018}. But again, exact formulas for the process $X$ are not mentioned and the Euler-scheme is used to get approximate transition densities of $X$. Here, we claim that under the hypothesis of isotropic movement, with diagonal diffusion matrix as in equation ~\ref{eq: RACVM equation}, both the velocity and the position processes have a closed form formula. \\

@@

\begin{figure}[H]
	\centering
	\begin{subfigure}{0.48\textwidth}
		\centering
		\includegraphics[scale=0.3]{images/rcvm/tau5_nu5_omega1mu11_sigmaobs0.03 .png}
		\caption{}
	\end{subfigure}
	\begin{subfigure}{0.48\textwidth}
		\centering
		\includegraphics[scale=0.3]{images/rcvm/hspeed_histo_tau5_nu5omega1_mu11_sigmaobs0.03 .png}
		\caption{}
	\end{subfigure}
	\caption{(a) Samples of a RACVM with parameters $\tau=5 h$, $\nu=5$ km/h,$\omega=1$ rad/h, $\mu=(1,0)$ km/h. The black cross indicates the starting point of the process. The samples were obtained from the exact transition densities. (b) Histogram of the empirical velocities obtained from the samples, along with the theoretical probability density function.}
	\label{fig: sample RACVM}
\end{figure}

\subsubsection{State-space model}
\label{section: state space model}
In most applications, it is needed to consider measurement errors through a state-space model as done by \cite{johnson_continuoustime_2008}. We suppose here that $n$ positions are observed at discrete times $t_j$, $j \in \{1,\cdots, n\}$. $Y_j$ is the random vector representing the observed position at time $t_j$, while $X_j=X(t_j)$ is the true position at time $t_j$. The statistical model is the following:

\begin{equation}  \left\{
	\begin{array}{l}
		Y_j=X_j+\varepsilon_j \quad j \in \{1,\cdots,n\}\\
		\varepsilon_j \underset{i.i.d}{\sim} \mathcal{N}(0,\sigma_{obs}^2)  \\
		dX(t)=V(t)dt  \\
		dV(t)=-A(V(t)-\mu)dt+\frac{2\nu}{\sqrt{\pi \tau}} dW(t) 
	\end{array}
	\right.
\end{equation}

We consider parameters $\tau$, $\nu$, $\mu$ and $\omega$ depending on time and make the approximation mentioned in \cite{michelot_varying-coefficient_2021} that these parameters are constant on each time step $[t_j,t_{j+1}]$, equal to their value at time $t_j$.  Denote $\Delta_j=t_{j+1}-t_j$, $V_j=V(t_j)$, $X_j=X(t_j)$, $\mu_j=\mu(t_j)$, $\tau_j=\tau(t_j)$, $\nu_j=\nu(t_j)$,$\omega_j=\omega(t_j)$ and $A_j=\begin{pmatrix} 
	\frac{1}{\tau_j} & -\omega_j \\
	\omega_j & \frac{1}{\tau_j}
\end{pmatrix}$.

\begin{equation}
	\forall j \in \{1, \cdots, n\}, \quad 
	\left\{ \begin{array}{l}
		V_{j+1}=\mu_j +\exp(-A_j\Delta_j)(V_j-\mu_j)+\zeta_j \\
		X_{j+1}=X_j+\mu_j \Delta_j+A_j^{-1}(I_2-\exp(-A_j\Delta_j))(_Vj-\mu_j)+\xi_j
	\end{array}\right. \\
	\label{eq: RACVM autoregressive formula}
\end{equation}

where $\zeta_j$ and $\xi_j$ are centered normal random vectors.
Denote $\alpha_j=\begin{pmatrix} X_{j,1}  & X_{j,2} & V_{j,1} & V_{j,2}\end{pmatrix}^\top$. The state space equations are

\begin{equation}
\left\{
\begin{array}{l}
	Y_j=Z\alpha_j+\varepsilon_j \\
	\alpha_{j+1}=T_j \alpha_j+B_j \mu_j + \eta_j
\end{array}
\right.
\label{eq: RACVM state space}
\end{equation}
where the link matrices are 
\[Z=\begin{pmatrix} I_2 & 0_{2,2}\end{pmatrix} \quad  T_j=\begin{pmatrix} I_2 & A_j^{-1}(I_2-\exp(-A_j \Delta_j)) \\ 0_{2,2} & \exp(-A_j \Delta_j) \end{pmatrix} \quad B_j=\begin{pmatrix}
	\Delta_j I_2-A_j^{-1}(I_2-\exp(-A_j\Delta_j)) \\
	I_2-\exp(-A_j\Delta_j)\end{pmatrix}
\]

$\varepsilon_j \sim \mathcal{N}(0,\sigma_{obs}^2 I_2)$ and the vector $\eta_j=\begin{pmatrix} \xi_{j,1} & \xi_{j,2} & \zeta_{j,1} & \zeta_{j,2} \end{pmatrix}^\top$ is a 4 dimensional normal random vector, with mean zero. Its covariance matrix $Q_j$ can be computed explicitly.
\begin{equation}
	Q_j=\begin{pmatrix}
		M_j & L_j \\
		L_j & N_j
	\end{pmatrix}
	\label{eq: state space covariance}
\end{equation}
where $M_j$, $N_j$and $L_j$ are block matrices of size $2 \times 2$ such that 
\[M_j=\left( \Delta-2 \frac{\omega_j \sin(\omega_j \Delta_j)-\frac{1}{\tau_j} \cos(\omega_j \Delta_j)}{\frac{1}{\tau_j^2}+\omega_j^2 } \exp\left( -\frac{\Delta_j}{\tau_j} \right) +\frac{\tau_j}{2} \left( \frac{\omega_j^2-\frac{3}{\tau_j^2}}{\frac{1}{\tau_j^2}+\omega_j^2}-\exp\left( -\frac{2\Delta_j}{\tau_j}\right)\right) \right)I_2\]

\[N_j=\frac{2\nu_j^2}{\pi}\left(1-e^{-\frac{2 \Delta_j}{\tau_j}}\right)I_2\]

\[L_j=\begin{pmatrix} l_1 & l_2 \\
	l_2 & l_1\end{pmatrix}\]
where 
\[l_1=\frac{4\nu_j^2}{2 \pi \tau_j \times \left( \frac{1}{\tau_j^2}+\omega_j^2\right)} \times \left( 1+\exp\left( -\frac{2\Delta_j}{\tau_j}\right)-2\exp\left( -\frac{\Delta_j}{\tau_j}\right)-2\exp\left( -\frac{\Delta_j}{\tau_j}\right) \cos(\omega_j \Delta_j)\right)\]
\[l_2=\frac{4\nu_j^2}{2 \pi \tau_j \times \left( \frac{1}{\tau_j^2}+\omega_j^2\right)} \times\left( \exp\left( -\frac{\Delta_j}{\tau_j}\right) \sin(\omega_j \Delta_j)-\frac{\omega_j \tau_j}{2} \left(1-\exp\left( -2 \frac{\Delta_j}{\tau_j}\right) \right)\right)\]

Again, one can check that for $\omega_j=0$, these formulas match the formulas of the state-space derived by \cite{johnson_continuoustime_2008} for the continuous-time correlated random walk model.
This state-space formulation is the basis for fast inference of time-varying parameters using the Kalman filter. This is the technique used in the \texttt{R} package \texttt{smoothSDE} \cite{michelot_varying-coefficient_2021} for the CVM, which corresponds to the special case $\omega=0$. Therefore, we extend this inference method to the more general model~\ref{eq: RACVM equation}.

@@

\subsection{Formulation of the model}

@@

\begin{figure}[H]
		\centering
		\includegraphics[scale=0.7]{images/simulation study/illustrative_sample_fjords.png}
	\caption{Sample of a CRCVM. x and y axis are in metres and initial positions are indicated by a red cross and red points represent the nearest points on the boundary of the domain. One  five day sample with time step $h=5$ min within the spatial constraints of Scoresby Sound fjords system in South-East Greenland. Parameters are $\nu=4$ km/h and $\tau=1$ h. }
	\label{fig: samples_fjords_CRCVM}
\end{figure}

@@

\subsubsection{Simulation study}
\label{section: simulation constrained motion}
In order to experiment the constrained models we simulated a CRCVM within different domains for smooth parameters $\omega$ depending on the angle $\Theta$ and the distance to the boundary of the domain $D^{shore}$.\\

100 batches of 6 individual trajectories over $12$ hours with time step $h=1$ min were sampled in a rectangular domain with sides of length $8$ km from a CRCVM. Gaussian noise with standard deviation $\sigma_{obs}=10$ m is considered to represent measurement error and approach a practical case. Denoting $n$ the number of observations, which is the same for each individual trajectory, $Y_{ij}$ the observation with noise and $X(t_{ij})=X_{ij}$ the true location, the precise statistical set up is the following:

\begin{equation}  \forall  i \in \{1,\cdots,6\}, \quad  \left\{
	\begin{array}{l}
		Y_{ij}=X_{ij}+\varepsilon_{ij}, \quad j \in \{1,\cdots, n\} \\
		\varepsilon_{ij} \underset{i.i.d}{\sim} \mathcal{N}(0,\sigma_{obs}^2) \\
		dX_i(t)=V_i(t) dt \quad \\
		dV_i(t)=-A(X_i(t),V_i(t))V_i(t)dt+\frac{2\nu_i}{\sqrt{\pi \tau_i}} dW(t)  \\
		\log(\tau_i)=\log(\tau_0)+b_{\tau,i}, \\\
		\log(\nu_i)=\log(\nu_0)+b_{\nu,i}, \\
		\begin{pmatrix} b_{\tau,i} \\ b_{\nu,i} \end{pmatrix} \underset{i.i.d}{\sim} \mathcal{N}\left(0,\begin{pmatrix} \sigma_{\tau}^2 & 0 \\ 0 & \sigma_{\nu}^2 \end{pmatrix}\right) \\
		\omega_i(t)=f_{\omega}(D^{shore}_i(t),\Theta_i(t)), \quad t \geq 0\\
		f_{\omega}(D^{shore},\Theta)=\omega_0+\sum_{k=1}^{L} \omega_k \psi_k(D_{shore},\Theta), \quad (D_{shore},\Theta) \in \R^+ \times [-\pi,\pi]
	\end{array}
	\right.
	\label{eq: rect simulation parameters}
\end{equation}
Here, $L=8$ and $\psi_k$ are bivariate basis functions constructed through tensor products of univariate cubic spline basis functions. We refer to \cite{wood_generalized_2017} section $4.
1.8$ for the construction of tensor splines basis function. We choose $\sigma_{\tau}=0.1$ h and $\sigma_{\nu}=0.3$ km/h to represent the individual variability, wheras $\tau_0$ and $\nu_0$ are fixed respectively to $1$ h and $4$ km/h. The values of the spline coefficients $\omega_k$, $k \in \{0,\cdots,L\}$ are given in Table~\ref{table: estimation performance rect}. For each trajectory, initial velocity is set to $0$ and initial position is uniformly sampled in the rectangular domain, at least $500$ metres away from the boundary. To simulate the trajectories, we suppose that the value of the parameter $\omega$ is constant on each time step, that is equal to 
\begin{equation}
	\omega_{ij}=f_{\omega}(D^{shore}_{ij},\Theta_{ij}), \quad j  \in \{1,\cdots,n\}
\end{equation}
  where $D^{shore}_{ij}=D^{shore}_i(t_{ij})$ and $\Theta_{ij}=\Theta_i(t_{ij})$. The next position and velocity are then obtained from equation~\ref{eq: RACVM state space}. None of the $600$ simulations reached the boundary. \\

100 batches of 6 individual trajectories over $24$ hours were also simulated in Scoresby Sound fjords system in which the narwhals move under constraints in our application. The parameters were set as in~\ref{eq: rect simulation parameters}, with only the coefficients $\omega_k$, $k \in \{1,\cdots,L\}$ changed to adapt to this specific domain. The exact values of the spline coefficients can be found in Table~\ref{table: estimation performance fjords}.
Again, for each trajectory, initial velocity was set to $0$ and initial position is uniformly sampled within the sea, with the requirement that it should be at least $500$ metres away from the shore. Time steps are also kept constant, equal to $1$ min. None of the $600$ trajectories reached land. Note that if time steps are too large, the assumption of constant covariates between observations is very likely to be violated due to the steepness of the shoreline, resulting in samples that reach land very often. For the fjords system, this already happens if we try to sample directly with $5$ min time steps. Moreover, sampling within the fjords takes much more time than within the rectangular domain, since it is necessary to loop over each polygon of the geometry to compute the nearest shore point for each new position.

@@

\subsection{Description of the movement data}

For each narwhal $i \in \{1,\cdots,N\}$ with $N=6$, we denote $n_i$ the total number of observations, which can be decomposed in $n_i=n_{i,pre}+n_{i,post}$ where $n_{i,pre}$ is the number of observations before exposure, and $n_{i,post}$ is the number of observations after exposure. We suppose that the position is observed at times $t_{i1}, t_{i2}, \cdots,t_{in_i}$ and  for $j \in \{1,\cdots,n_i\}$, we denote $y_{ij}=\begin{pmatrix} y_{ij1} & y_{ij2} \end{pmatrix}^\top$ the observed GPS position at time $t_{ij}$, converted in Easting and Northing.

>>>>

In Section 2, we give an overview of the narwhal data available for the analysis.
The diffusion models are discussed in Section 3. Inference and simulation study results are explained in section 4.
Then, our models are applied to analyse a behavioral response of the narwhals in section 5. The last section is dedicated to a discussion about these results, and potential extension of the models.

\section{Movement data of six narwals in South-East Greenland}

\subsection{Description of the controlled exposure experiments}

For each narwhal $i \in \{1,\cdots,N\}$ with $N=6$, we denote $n_i$ the total number of observations, which can be decomposed in $n_i=n_{i,pre}+n_{i,post}$ where $n_{i,pre}$ is the number of observations before exposure, and $n_{i,post}$ is the number of observations after exposure. We suppose that the positions are observed at times $t_{i1}, t_{i2}, \cdots,t_{in_i}$ and  for $j \in \{1,\cdots,n_i\}$, we denote $y_{ij}=\begin{pmatrix} y_{ij1} & y_{ij2} \end{pmatrix}^\top$ the observed GPS position at time $t_{ij}$, converted in Easting and Northing. These are noisy observations of the underlying unobserved true positions $x_{ij}$.\\

The land geometry for the specific region in South-East Greenland is gathered from \href{https://www.openstreetmap.org/#map=11/70.4029/-27.2928}{OpenStreetMap} database. The land polygons define the boundaries of the domain $\mathcal{D}$ in which the narwhals can move. The closest point on the shoreline to the observed GPS position $y_{ij}$  is denoted $\hat{p}_{ij}$. Due to GPS measurement errors, it is an approximation of the unobserved closest point $p_{ij}$ to the true position, and therefore the distance to $\hat{p}_{ij}$, denoted $\hat{D}^{shore}_{ij}$ in the sequel, is considered a close approximation of the actual distance to the boundary $D^{shore}_{ij}$. The resulting values range from $0$ to $7.5$ km. About $9\%$ of the GPS measurements in the dataset turned out to be on land. This is primarily attributed to inaccuracies in the shoreline maps rather than errors in GPS measurements. For constrained movement models, these points are ignored. Then, exposure to the shore is defined with the following formula
\begin{equation}
	\left\{
	\begin{array}{ll}
		E^{shore}_{ij}=0 & \mbox{if } D^{shore}_{ij}>D_{ref} \\
		E^{shore}_{ij}=\frac{1}{D_{shore}(t_j)} & \mbox{if } D^{shore}_{ij} \leq D_{ref}
	\end{array}
	\right.
	\label{eq: exp shore definition}
\end{equation}
where the reference threshold $D_{ref}$ is chosen equal to $0.5$ km. We denote $\hat{E}^{shore}_{ij}$ the approximation of $E^{shore}_{ij}$ obtained by replacing the true distance $D^{shore}_{ij}$ by $\hat{D}^{shore}_{ij}$.
The angle between the narwhal's heading and the shoreline, denoted $\Theta_{ij}$, is estimated as the angle $\hat{\Theta}_{ij}$ between the observed empirical velocity $\hat{v}_{ij}=\frac{y_{i,j+1}-y_{ij}}{\Delta_j}$, where $\Delta_j=t_{i,j+1}-t_{ij}$, and the vector $n_{ij}=y_{ij}-p_{ij}$. This is illustrated in figure~\ref{fig: illustration nearest shore points}. A value $\Theta_{ij}=\pm \frac{\pi}{2}$ indicates movement parallel to the shore, while $\Theta_{ij} \in \left]\frac{\pi}{2},\pi\right] \cup \left[-\pi,-\frac{\pi}{2}\right[$ indicates movement toward the shore.

		%nodes
		\node[left] at (-1,3.5) {Land};
		\node[right] at (1,3.5) {Water};
		\node[left] at (0,0) {$\partial \mathcal{D}$};
		\node[left] at (0,2) {$p_{ij}$};
		\node[left] at (1.2,2.8) {$\hat{v}_{ij}$};
		\node[below] at (1.8,2) {$y_{ij}$};
		\node[below,red] at (3,2) {$\vec{n}_{ij}$};

The distance from the ship, denoted $D^{ship}_{ij}$, is computed from the GPS ship and narwhals's positions. The values are comprised between $2.68$ and $63.8$ km.
Exposure to ship covariate for narwhal $i$ at time $t_{ij}$, denoted $E^{ship}_{ij}$, is defined as the inverse of the distance to the ship (in km) \cite{heide-jorgensen_behavioral_2021}.
$D^{ship}$ is only known when the narwhal is in line of sight with the boat, so that exposures are set to $0$ when this is not this case. This implies that the effect of the ship noise on the narwhals's movement is only considered in this specific case in the whole study, though it is very likely that narwhals can perceive the disturbance even when they are not clearly in line of sight with the sound source. This provides a conservative estimate of the effect of the noise exposure. The covariate $E^{ship}$ is meant to be a proxy for sound exposure levels received by the narwhals. Reasonably, the closer the narwhal is to the ship, the more it can hear its sound, and the greater the exposure. Exposure levels for each narwhals are displayed in figure~\ref{fig: realexpthroughtime}.

@@

@@

\begin{equation*} \left\{
	\begin{array}{l}
		dX(t)=V(t)dt \\
		dV(t)=-A(V(t)-\mu)dt+\frac{2\nu}{\sqrt{\pi \tau}} dW(t) 
	\end{array}
	\right.
	\label{eq: RACVM equation bis}
\end{equation*}
where $X$ is the two-dimensional location and $V$ is the two-dimensional velocity.
In this section, we obtain explicit distributional results about the process $X$. Though this model is considered in several papers, we found no mention of the closed form formulas derived here. In \cite{gurarie_correlated_2017}, only the process $V$ is exhaustively studied, and the classical distributional results about this process are used for estimation and simulation purposes.
A more general formulation is considered by \cite{albertsen_generalizing_2018} in which the diffusion matrix is lower triangular matrix with positive diagonal elements whereas and two distinct autocorrelation parameters are considered in both directions allowing for anisotropic movement.
In this case, \cite{albertsen_generalizing_2018} derives an explicit form for the gaussian transition density of the velocity process, where the covariance matrix is expressed using a Kronecker sum \cite{albertsen_generalizing_2018}. But again, exact formulas for the process $X$ are not mentioned and the Euler-scheme is used to get approximate transition densities of $X$. Here, we claim that under the hypothesis of isotropic movement, with diagonal diffusion matrix as in equation ~\ref{eq: RACVM equation}, not only the velocity but also the position process has a closed form formula. \\

@@

\begin{figure}[H]
	\centering
	\begin{subfigure}{0.48\textwidth}
		\centering
		\includegraphics[scale=0.3]{images/illustrative_sample_standard.png}
		\caption{}
	\end{subfigure}
	\begin{subfigure}{0.48\textwidth}
		\centering
		\includegraphics[scale=0.3]{images/illustrative_sample_fjords.png}
		\caption{}
	\end{subfigure}
	\caption{(a) Samples of a RACVM with parameters $\tau=1 h$, $\nu=4$ km/h,$\omega=0.5$ rad/h, $\mu=(0,0)$ km/h and constant time step $h=5$ min. The red cross indicates the starting point of the process. The samples were obtained from the exact transition densities. (b) Samples of a CRCVM with parameters $\tau=1 h$, $\nu=4$ km/h,$\omega$ as a smooth function of covariates $\Theta$ and $D^{shore}$ as described in section~\ref{section: CRCVM}. Position are sampled from constant time step $h= 1$ min and then doownsampled to $5$ min.The samples were obtained from approximate transition densities  with constant covariates on each time step. }
	\label{fig: sample RACVM}
\end{figure}

@@

@@

@@

\section{Inference}

For the baseline model, only the effect of the shore is included through the shore exposure covariate $E^{shore}$ and the angle covariate $\Theta$.\\
Although individuals have a tendency to move in the same way, their movement don't have exactly the same characteristics depending on their age and size for instance. This is taken into account by using a mixed effect model for the RCVM parameters.
For each narwhal $i \in \{1, \cdots, N\}$, we have 	GPS observations $Y_{ij}$ at times $t_{ij}$, $j \in \{1,\cdots,n_{i,pre}\}$ and the true motion processes $V_i$ and $X_i$ are supposed to solve equation~\ref{eq: CRCVM equation}. Because $\tau$ and $\nu$ can only be positive, a log link function is used for these parameters. The angular velocity $\omega$ is expressed as a smooth function of $\Theta$ and $E^{shore}$ through splines. Since both covariates are not on the same scale (one is in $rad$ while the other is in $km^{-1}$), tensor splines were judged more appropriate than thin plate regression splines \cite{wood_generalized_2017}.\\

Denoting $n_{i,pre}$ the number of observations before exposure for the narwhal $i$, and $\tau_i$, $\nu_i$, $\omega_i$ the parameters of the diffusion .

\begin{equation}  \forall i \in \{1,\cdots,N\}, \quad \left\{
	\begin{array}{l}
		Y_{ij}=X_{ij}+\varepsilon_{ij} \quad j \in \{1,\cdots,n_{i,pre}\}\\
		\varepsilon_{ij} \underset{i.i.d}{\sim} \mathcal{N}(0,\sigma_{obs}^2)  \\
		dX_i(t)=V_i(t)dt  \\
		dV_i(t)=-\begin{pmatrix} 
			\frac{1}{\tau_i} & -\omega_i(t) \\
			\omega_i(t) & \frac{1}{\tau_i}
		\end{pmatrix}V_i(t)dt+\frac{2\nu_i}{\sqrt{\pi \tau_i}} dW(t) \\
		\log(\tau_i)=\log(\tau_{pre})+b_{\tau,pre}^{(i)} \\
		\log(\nu_i)=\log(\nu_{pre})+b_{\nu,pre}^{(i)}  \\
		\begin{pmatrix} b_{\tau,pre}^{(i)} \\ b_{\nu,pre}^{(i)} \end{pmatrix} \underset{i.i.d}{\sim} \mathcal{N}\left(0,\begin{pmatrix} \sigma_{\tau,pre}^2 & 0 \\ 0 & \sigma_{\nu,pre}^2 \end{pmatrix}\right) \\
		\omega_i(t)=f_{\omega}(E^{shore}_i(t),\hat{\Theta}_i(t))\\
		f_{\omega}(E^{shore},\Theta)=\omega_{0}+\sum_{k=1}^{L} \omega_{k} \psi_k(E^{shore},\Theta), \quad (E^{shore},\Theta) \in \R^+ \times [-\pi,\pi]
	\end{array}
	\right.
	\label{eq: exact baseline model}
\end{equation}

$\log(\tau_{pre})$, $\log(\nu_{pre})$, $\omega_{0,pre}$ are population intercepts and $b^{(i)}_{\tau,pre}$, $b^{(i)}_{\nu,pre}$
are the individual parameters.

The functions $\psi_k$ in equation~\ref{eq: exact baseline model} are bivariate basis functions constructed through tensor products of univariate cubic spline basis functions. We refer to \cite{wood_generalized_2017} section $4.
1.8$ for the construction of tensor splines basis functions.

%Add plot of the spline basis functions \\
The splines degree of freedom is $L=q_E \times q_{\Theta}$ and the numbers of knots in the splines is $q_E-1$ and $q_{\Theta}-1$ for the respective covariates $E_{shore}$ and $\Theta$. It is needed to choose them before hand. One possibility to help choosing the best value is to increase the degree of freedom step by step  and find the value for which the decrease in AIC starts to get lower.\\

\begin{equation}  \forall i \in \{1,\cdots,N\}, \quad \left\{
	\begin{array}{l}
		Y_{ij}=X_{ij}+\varepsilon_{ij} \quad j \in \{n_{i,pre}+1,\cdots,n_{i,post}\}\\
		\varepsilon_{ij} \underset{i.i.d}{\sim} \mathcal{N}(0,\sigma_{obs}^2)  \\
		dX_i(t)=V_i(t)dt  \\
		dV_i(t)=-\begin{pmatrix} 
			\frac{1}{\tau_i} & -\omega_i(t) \\
			\omega_i(t) & \frac{1}{\tau_i}
		\end{pmatrix}V_i(t)dt+\frac{2\nu_i}{\sqrt{\pi \tau_i}} dW(t) \\
		\log(\tau_{ij})=\log(\tau_{0,post}) +\sum_{k=1}^{q} \log(\tau_{k,post})\phi_{k}(E^{ship}_i(t))+b^{(i)}_{\tau,post} \\
		\log(\nu_{ij})=\log(\nu_{0,post}) +  \sum_{k=1}^{q} \log(\nu_{k,post}) \phi_{k}(E^{ship}_i(t)) +b^{(i)}_{\nu,post}  \\
		\omega_i(t)=f_{\omega}(E^{shore}_i(t),\Theta_i(t))\\
		f_{\omega}(E^{shore},\Theta)=\omega_{0}+\sum_{k=1}^{L} \omega_{k} \psi_k(E^{shore}_i,\Theta_i), \quad (E^{shore},\Theta) \in \R^+ \times [-\pi,\pi]
	\end{array}
	\right.
	\label{eq: exact response model}
\end{equation}

Again, $\log(\tau_{0,post})$, $\log(\nu_{0,post})$, $\omega_{0,post}$ are population intercepts, $b^{(i)}_{\tau,post}$,$b^{(i)}_{\nu,post}$ are individual parameters. The functions $\phi_k$ are cubic spline basis functions and $q$ is the spline degree of freedom for the exposure to the ship covariate.

\subsection{State-space model}
\label{section: state space model}
In most applications, it is needed to consider measurement errors through a state-space model as done by \cite{johnson_continuoustime_2008}. This is the approach we use here. In the statistical model~\ref{eq: exact baseline model} for the baseline model, we make the approximation mentioned in \cite{michelot_varying-coefficient_2021} that the parameter $\omega_i$ is constant on each time step $[t_{ij},t_{i,j+1}]$, equal to its value at time $t_{ij}$.\\
 Denote $\Delta_{ij}=t_{i,j+1}-t_{ij}$, $V_{ij}=V(t_{ij})$, $X_{ij}=X(t_{ij})$, ,$\omega_{ij}=\omega(t_{ij})$, $A_{ij}=\begin{pmatrix} 
	\frac{1}{\tau_{i}} & -\omega_{ij} \\
	\omega_{ij} & \frac{1}{\tau_{i}}
\end{pmatrix}$ and $\alpha_{ij}=\begin{pmatrix} X_{ij,1}  & X_{ij,2} & V_{ij,1} & V_{ij,2}\end{pmatrix}^\top$.
\begin{comment}
\begin{equation}
	\forall j \in \{1, \cdots, n_{i,pre}\}, \quad 
	\left\{ \begin{array}{l}
		V_{i,j+1}=\exp(-A_{ij}\Delta_{ij})V_{ij}+\zeta_{ij} \\
		X_{i,j+1}=X_{ij}+A_{ij}^{-1}(I_2-\exp(-A_{ij}\Delta_{ij}))V_{ij}+\xi_{ij}
	\end{array}\right. \\
	\label{eq: RACVM autoregressive formula}
\end{equation}

where $\zeta_{ij}$ and $\xi_{ij}$ are centered normal random vectors.
\end{comment}

The state space equations are

\begin{equation}
	\left\{
	\begin{array}{l}
		Y_{ij}=Z\alpha_{ij}+\varepsilon_{ij} \\
		\alpha_{i,j+1}=T_{ij} \alpha_{ij} + \eta_{ij}
	\end{array}
	\right.
	\label{eq: RACVM state space}
\end{equation}
where the link matrices are 
\[Z=\begin{pmatrix} I_2 & 0_{2,2}\end{pmatrix} \quad  T_{ij}=\begin{pmatrix} I_2 & A_{ij}^{-1}(I_2-\exp(-A_{ij} \Delta_j)) \\ 0_{2,2} & \exp(-A_{ij} \Delta_j) \end{pmatrix}\]

The vector $\eta_{ij}=\begin{pmatrix} \xi_{ij,1} & \xi_{ij,2} & \zeta_{ij,1} & \zeta_{ij,2} \end{pmatrix}^\top$ is a 4 dimensional normal random vector, with mean zero. Its covariance matrix $Q_{ij}$ can be computed explicitly.
\begin{equation}
	Q_{ij}=\begin{pmatrix}
		M_{ij} & L_{ij} \\
		L_{ij} & N_{ij}
	\end{pmatrix}
	\label{eq: state space covariance}
\end{equation}
where $M_{ij}$, $N_{ij}$and $L_{ij}$ are block matrices of size $2 \times 2$ such that 
\[M_{ij}=\frac{\sigma_i^2}{C_{ij}}\left( \Delta-2 \frac{\omega_{ij} \sin(\omega_{ij} \Delta_{ij})-\frac{1}{\tau_{i}} \cos(\omega_{ij} \Delta_{ij})}{C_{ij} } \exp\left( -\frac{\Delta_{ij}}{\tau_{ij}} \right) +\frac{\tau_{i}}{2} \left( \frac{\omega_{ij}^2-\frac{3}{\tau_{ij}^2}}{C_{ij}}-\exp\left( -\frac{2\Delta_{ij}}{\tau_i}\right)\right) \right)I_2\]

\[N_{ij}=\frac{\sigma_i^2}{2}\left(1-e^{-\frac{2 \Delta_{ij}}{\tau_{i}}}\right)I_2\]

\[L_{ij}=\begin{pmatrix} a & b \\
	b & a\end{pmatrix}\]
with
\[a=\frac{\sigma_i^2}{2 C_{ij}^2 } \times \left( 1+\exp\left( -\frac{2\Delta_{ij}}{\tau_{i}}\right)-2\exp\left( -\frac{\Delta_{ij}}{\tau_{i}}\right)-2\exp\left( -\frac{\Delta_{ij}}{\tau_{i}}\right) \cos(\omega_{ij}\Delta_{ij})\right)\]
\[b=\frac{\sigma_{i}^2}{2 C_{ij}^2} \times\left( \exp\left( -\frac{\Delta_{ij}}{\tau_{i}}\right) \sin(\omega_{ij} \Delta_{ij})-\frac{\omega_{ij} \tau_{i}}{2} \left(1-\exp\left( -2 \frac{\Delta_{ij}}{\tau_{i}}\right) \right)\right)\]
and $C_{ij}=\frac{1}{\tau_i^2}+\omega_{ij}^2$, $\sigma_{i}=\frac{2\nu_i}{\sqrt{\pi \tau_i}}$.
Again, one can check that for $\omega_{ij}=0$, these formulas match the formulas of the state-space derived by \cite{johnson_continuoustime_2008} for the continuous-time correlated random walk model.
This state-space formulation is the basis for fast inference of time-varying parameters using the Kalman filter. This is the technique used in the \texttt{R} package \texttt{smoothSDE} \cite{michelot_varying-coefficient_2021} for the CVM, which corresponds to the special case $\omega=0$. Therefore, we extend this inference method to the more general model~\ref{eq: CRCVM equation} and use it for estimation of the baseline and the response models.

\section{Simulation study for the baseline model}
\label{section: simulation constrained motion}
In order to experiment the constrained models we simulated a CRCVM within different domains for smooth parameters $\omega$ depending on the angle $\Theta$ and the distance to the boundary of the domain $D^{shore}$. We considered observation of a baseline model as described in equation~\ref{eq: exact baseline model}\\.

We simulate $M=100$ batches of $N=6$ individual trajectories over $T=12$ h with time step $h=1$ min in a rectangular domain with sides of length $8$ km from a CRCVM. This amounts to $n_{i,pre}=720$ observations for each individual $i \in \{1,\cdots,N\}$. We fix $\sigma_{obs}=5$ m to represent low measurement errors. 
We choose $L=8$ for the degree of freedom of the bivariate splines of $D^{shore}$ and $\Theta$. The random effects standard deviations are set to $\sigma_{\tau,pre}=0.1$ h and $\sigma_{\nu,pre}=0.1$ km/h to represent the individual variability, and $\tau_{pre}$ and $\nu_{pre}$ are fixed respectively to $1$ h and $4$ km/h. The exact values of the spline coefficients $\omega_k$, $k \in \{0,\cdots,L\}$ used to simulate the trajectories are given in Table~\ref{table: estimation performance rect}. For each trajectory, initial velocity is set to $(0,0)$ and initial position is uniformly sampled in the rectangular domain, at least $500$ metres away from the boundary. WE kept the same $6$ initial positions for all the batches of trajectories. Toobtain teh samples of the CRCVM, we suppose that the value of the parameter $\omega$ is constant on each time step, that is equal to 
\begin{equation}
	\omega_{ij}=f_{\omega}(D^{shore}_{ij},\Theta_{ij}), \quad j  \in \{1,\cdots,n\}
\end{equation}
where $D^{shore}_{ij}=D^{shore}_i(t_{ij})$ and $\Theta_{ij}=\Theta_i(t_{ij})$. The next position and velocity are then obtained from equation~\ref{eq: RACVM state space}. None of the $600$ simulations reached the boundary. \\

We also simulate $M=100$ batches of $N=6$ individual trajectories over $T=12$ with time step $h=1$ min in Scoresby Sound fjords system in which the narwhals move under constraints in our application. The parameters $\sigma_{obs}$,$L$,$\sigma_{\tau,pre}$,  $\sigma_{\nu,pre}$ , $\tau_{pre}$,$\nu_{pre}$ are kept the same as in the recangular case, and only the coefficients $\omega_k$, $k \in \{1,\cdots,L\}$ are changed to adapt to this specific fjords domain. The exact values of the spline coefficients are found in Table~\ref{table: estimation performance fjords}.
Again, for each trajectory, initial velocity is set to $0$ and initial position is uniformly sampled within the sea, with the requirement that it should be at least $500$ metres away from the shore.  None of the $M\times N=600$ trajectories reached land.

\begin{comment}Note that if time steps are too large, the assumption of constant covariates between observations is very likely to be violated due to the steepness of the shoreline, resulting in samples that reach land very often. For the fjords system, this already happens if we try to sample directly with $5$ min time steps. Moreover, sampling within the fjords takes much more time than within the rectangular domain, since it is necessary to loop over each polygon of the geometry to compute the nearest shore point for each new position. 
\end{comment}

@@
